---
title: "Week 13b: SQL"
output:
  revealjs::revealjs_presentation:
    theme: white
    center: true
    transition: none
    incremental: true
---


```{r setup, include=FALSE}
library(dplyr)
library(knitr)
library(rvest)
library(stringr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = T)
```

# SQL

## SQL Overview

- Generally SQL servers are created to host a large amound of data. 
- For this class we will use SQLite which enables users to store database files locally, but the principles are the same for querying a server-based database.

## Baseball Data

We will use the baseball database  [https://www.kaggle.com/seanlahman/the-history-of-baseball](https://www.kaggle.com/seanlahman/the-history-of-baseball). Recall we used this earlier in the class with `dplyr` commands. There is a SQLite database file that you can directly access, but instead we will build our own SQLite database.

## Building Database: 1. Read in Data

```{r}
library(readr)
library(dplyr)
player <- read_csv('https://raw.githubusercontent.com/stat408/Lecture12b/master/player.csv')
batting <- read_csv('https://raw.githubusercontent.com/stat408/Lecture12b/master/batting.csv')

batting <- batting %>% select(-sf, -g_idp)

```


## Building Database: 2. Establish Database
```{r}
library(RSQLite)

db_connection <- dbConnect(SQLite(), "practice_DB.db")

```

## Building Database: 3. Add Tables
```{r, eval = T}
dbWriteTable(db_connection, "batting", batting, overwrite = T)
dbWriteTable(db_connection, "player", player, overwrite = T)
dbListTables(db_connection)

```



## Building Database: 4. Fields in Tables
```{r}
dbListFields(db_connection, "batting")
dbDisconnect(db_connection)
```

# SQL commands

## SQL commands

The most basic SQL queries have the following structure:

- **SELECT** var1name, var2name (filter columns)
- **FROM** tablename (identify table)
- **WHERE** condition1 (filter rows)
- **GROUP_BY** var3name (aggregate data)
- **HAVING** condition2 (filter aggregated data)
- **ORDER_BY** var (arrange ordering)


## SQL commands

SQL also allows join operators: 

- **INNER JOIN** 
- **OUTER JOIN** 
- **LEFT JOIN**
- **FULL JOIN** 



## SQL Query 1
Select all columns for player
```{r}
db_connection <- dbConnect(SQLite(), "practice_DB.db")
dbGetQuery(db_connection,"SELECT * 
                          FROM Player")
```

## SQL Query 2
Select name_first, name_last for all players for your birthday
```{r}
dbGetQuery(db_connection,
"SELECT name_first, name_last
FROM player 
WHERE birth_month = 10 AND birth_day = 15")
```

## SQL Query 3
Compute average weight for players born on your birthday 
```{r}
dbGetQuery(db_connection,
"SELECT AVG(weight) 
FROM player 
WHERE birth_month = 10 AND birth_day = 15")
```

## SQL Query 4: Part 1
Compute the total number of homeruns hit for players born on your birthday 
```{r}
dbExecute(db_connection,
"CREATE TABLE birthday as
SELECT name_first, name_last, player_id
FROM player 
WHERE birth_month = 10 AND birth_day = 15")

dbListTables(db_connection)
```



## SQL Query 4: Part 2
Compute the total number of homeruns hit for players born on your birthday 

```{r}
dbGetQuery(db_connection,
"SELECT SUM(hr)
FROM birthday as bd
LEFT JOIN batting as b
ON bd.player_id = b.player_id
")
```


